Optimal Light Curve Attitude Inversion with Measurement Noise: Two Case Studies

### 2025-03-25 paper.tex
submitted

% \makeatletter
% \AtBeginDocument{%
%   \expandafter\renewcommand\expandafter\subsubsection\expandafter{%
%     \expandafter\FloatBarrier\subsubsection
%   }%
% }
% \makeatother

\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{times}
\usepackage[numbers]{natbib}
\usepackage{graphicx}
\usepackage{bm} % for bold math
\usepackage{mathtools} % for \prescript
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{adjustbox}

@@

\newcommand{\figbig}[0]{0.5\textwidth}
\newcommand{\figmed}[0]{0.4\textwidth}
\newcommand{\figsmall}[0]{0.3\textwidth}

\newcommand{\subfigmed}[0]{0.4\textwidth}
\newcommand{\subfigsm}[0]{0.29\textwidth}
\newcommand{\subfigbig}[0]{0.43\textwidth}

@@

Methods for estimating the attitude state of uncontrolled space objects from light curves differ significantly depending on the presence of an initial guess. If prior knowledge is available, many filter-based approaches have been developed to process new photometry to update the guess orientation and angular velocity. This class of methods includes single Kalman filters \cite{burton2021two, gagnon2024, wetterer2009}, and multi-filter multiple-model adaptive estimation (MMAE) schemes \cite{linares2014space, dianetti2020}. These approaches can be very effective if the initial state estimate is close to the truth, but have been reported to diverge even in cases when the initial guess is within $5^\circ$ of the ground truth orientation and $0.03^\circ/s^{-1}$ in angular velocity error \cite{gagnon2024} due to measurement nonlinearity and ambiguities.

@@

While much work has performed full state attitude inversion on simulated light curve data \cite{burton2024journal, gagnon2024, robinson2025att}, studies using real data often focus on estimating an assumed constant spin rate and axis of rotation for the observed object. These assumptions simplifies the estimation process, but can lead to misleading conclusions when the observed object has a complex rotation state.

@@

The simplifying assumptions of constant, single-axis rotation that are commonly made when working with real data enable elegant solutions, but break down when the observed object is in a complex spin profile. Methods for full attitude state solutions -- time-varying orientation and angular velocities -- have been developed \cite{clark2022, burton2024journal, gagnon2024, linares2014particle}, but are seldom applied to real data. Works that have been successful at estimating a full rigid body attitude state from real data include genetic algorithms \cite{piergentili2020, gallucci2020} and brute-force grid search \cite{shafer2017}. Any full-state estimation method must tackle the inherent computational complexity, lack of knowledge of material reflectivities, and the presence of measurement noise.

In this paper, we describe a global full-state attitude solver inspired by recent PSO work \cite{burton2024journal} but is easier to tune, can be fully parallelized, and is robust to significant measurement noise \cite{robinson2025att}. Further, our method differs from other global estimation methods in that we do not seek to find one optimal solution and instead search for a collection of high-quality local minima to directly study the many ambiguous solutions for a given light curve. Our approach is designed specifically for use with real data as we take time-varying measurement noise and uncertainties in the body moments of inertia into account explicitly.

@@

This work proceeds by describing the inversion method, discussing efficient methods for computing the time-varying probability density of the light curve. We present inversion results for a simulated rocket body light curve where the ground truth is available to demonstrate the efficacy of the method, followed by results for real light curve data collected by the Purdue Optical Ground Station for a derelict ECHOSTAR satellite and Delta I rocket body. A deeper treatment of our method and additional simulated light curve results can be found in \cite{robinson2025att}.

@@

@@

\begin{equation} \label{eq:nll_loss}
 f(\vctr{x}) = \frac{1}{m}\sum_{k=1}^{m}\left[\frac{1}{2} \ln2\pi + \ln\sigma_k + \frac{1}{2}\left(\frac{\vctr{S}_k - \hat{\vctr{S}}_k(\vctr{x})}{\sigma_k}\right)^2 \right].
 \end{equation}

@@

In this work, initial state vectors are sampled from a prior distribution, usually defined independently for the MRP, angular velocity, and inertia tensor ratios. If no information is available, as is usually the case, a uniform distribution is sampled from for the first six components of the state. In particular, $\vctr{p}_{i,0}(0)$ is uniformly randomly sampled from orientation space, $\vctr{\omega}_{i,0}(0)$ is randomly sampled from a uniform distribution of magnitudes up to $\norm{\vctr{\omega}}_\text{max}$ -- the maximum angular rate at which to initialize state samples. We determine $\norm{\vctr{\omega}}_\text{max}$ based on the maximum power positive frequency in the observed light curve $f^\star$ in Hertz and an arbitrary safety factor $f_\omega \geq 1$, selected as $f_\omega = 1.25$ for this work:

\begin{equation} \label{eq:ang_vel_max}
  \norm{\vctr{\omega}}_\text{max} = 2\pi f_\omega f^\star.
\end{equation}

Uniformly sampling over angular velocity directions, we can express the default sampling strategy for an angular velocity state as:

@@

\noindent
where $g_{1:3} \in \mathcal{N}(0,1)$ and $m=\mathcal{U}(c_1 \norm{\vctr{\omega}}_\text{max}, c_2 \norm{\vctr{\omega}}_\text{max})$. The constants $c_1<1$ and $c_2>1$ are chosen to correct for the fact that even for a well-chosen bound $\norm{\vctr{\omega}}_\text{max}$, there can still be numerous high-quality solutions at significantly lower or higher angular velocities that possess equal $f^\star$ values. This is a result of object shape symmetries or harmonics in spin and precession rates. We select $c_1=0.5$, $c_2=2$ for this work.

@@

\begin{equation}
  \begin{bmatrix}
    \left(J_y / J_x\right)_{i,0} \\ \left(J_z / J_x\right)_{i,0}
  \end{bmatrix} \sim \mathcal{N} \left(\begin{bmatrix}
    \left(J_y / J_x\right)_\text{est} \\
    \left(J_z / J_x\right)_\text{est} \end{bmatrix}, \sigma_J I_2 \right),
\end{equation}

@@

\subsection{Generating the Observed Light Curve}

In order to make this simulated case realistic, we use the presented measurement noise models and introduce a mismatch between the ground truth object used to generate the measurements and the model used in inversion. The selected object is a high-fidelity model of the Saturn V second stage scaled down by a factor $10/3$ to act as a template rocket body object, shown on the right of Figure \ref{fig:rb_models}

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=\figsmall]{rocket_bodies2.png}
%   \caption{Simulated truth rocket body model \cite{nasa_models}.}
%   \label{fig:meas_model}
% \end{figure}

@@

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Simulated observation characteristics}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    Target COSPAR ID & 1990-054D \\ \grule
    First obs.\ (UTC) & Mar 8, 2025 02:00:00.000 \\ \grule
    Light curve duration & $5$ minutes \\ \grule
    Observations & $100$ \\ \grule
    Integration time & $0.5$ seconds \\ \brule
  \end{tabular}
  \label{tb:case1_in}
\end{table}

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Reflection properties of materials used in the synthetic data test case. The white paint parameters are estimated qualitatively.}
  \vspace*{6pt}
  \begin{tabular}{@{} l l l l @{}}
    \toprule
    Material & $C_d$ & $C_s$ & $n$ \\ \midrule
    Aluminum \cite{fankhauser2023} & $0.34$ & $0.40$ & $8.9$ \\ \arrayrulecolor{darkgray}\cmidrule(){1-4}
    White paint & $0.9$ & $0.1$ & $1$ \\ \brule
  \end{tabular}
  \label{tb:synth_matprops}
\end{table}

@@

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{True object attitude profile for all test cases}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    Initial $\vctr{p}(0)$ & $-\frac{1}{3} \begin{bmatrix} 1 & 1 & 1 \end{bmatrix}^T$ \\ \grule
    True $\vctr{\omega}(0)$ ($\text{rad}\cdot\text{s}^{-1}$) & $\vctr{\omega}(0) = \begin{bmatrix} 0.03 & 0.06 & 0.03 \end{bmatrix}^T$ \\ \grule
    Inertia tensor ratios & $J_y / J_x = 1, \: J_z / J_x = 0.25$ \\ \grule
    External torque $\vctr{M}$ & $\begin{bmatrix} 0 & 0 & 0 \end{bmatrix}^T$ \\ \brule
  \end{tabular}
  \label{tb:synth_att}
\end{table}
\FloatBarrier

@@

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=\figmed]{light_curves_adu_rocket_body_sdc.png}
%   \caption{Simulated observed flux for the rocket body.}
%   \label{fig:obs_adu_synth}
% \end{figure}

@@

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=\figsmall]{light_curves_std_rocket_body_sdc.png}
%   \caption{Simulated observed light curve standard deviation computed with Equation \ref{eq:sigma_total}.}
%   \label{fig:obs_std_synth}
% \end{figure}

\subsection{Case 1a and 1b: Synthetic Light Curve Optimization Results}

@@

Two distinct inversion cases are computed. Case 1a presents inversion results given no attitude information besides an estimate of the inertia tensor ratios derived from the known geometry, while Case 1b presents results given good prior attitude state distributions. Tables \ref{tb:case1a_prior} and \ref{tb:case1b_prior} summarize the initial condition sampling strategies employed for each case.

For both cases, a total of $n_\text{sample}=10^5$ initial state samplees are created via the sampling schemes described their respective tables. BFGS minimization is then performed for each initial state sample to locally solve the optimization problem given by Equation \ref{eq:opt_problem}.

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Initial condition sampling parameters used for inversion in Case 1a}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    $\vctr{p}$ sampling & Uniform, via Equation \ref{eq:mrp_sampler} \\ \grule
    $\vctr{\omega}$ sampling & Uniform, via Equation \ref{eq:omega_sampler} \\ \grule
    $J$ sampling & Gaussian, via Equation \ref{eq:omega_sampler} \\ \grule
    $\sigma_J$ & $0.1$ $[\text{nondim}]$ \\ \grule
    $\norm{\vctr{\omega}}_\text{max}$ & $5.26$ $[\text{deg}\cdot \text{s}^{-1}]$ \\ \brule
  \end{tabular}
  \label{tb:case1a_prior}
\end{table}

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Informative prior parameters used for inversion in Case 1b. The standard deviations $\sigma_\omega$ and $\sigma_J$ are for spherically-symmetric Gaussians centered on the ground truth. MRPs are sampled by rotating by $\theta\sim \mathcal{N}(0,\sigma_p)$ degrees away from the ground truth orientation around an axis uniformly sampled on the sphere.}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    $\vctr{p}$ sampling & Gaussian, $\sigma_p=20$ $[\text{deg}]$ \\ \grule
    $\vctr{\omega}$ sampling & Gaussian, $\sigma_\omega = 0.1$ $[\text{deg}\cdot\text{s}^{-1}]$ \\ \grule
    $J$ sampling & Gaussian, $\sigma_J = 0.05$ $[\text{nondim}]$ \\ \brule
  \end{tabular}
  \label{tb:case1b_prior}
\end{table}

Solutions are then classified as candidates for the ground truth with $\ell = 1/2$ via Equation \ref{eq:ell_selection_criteria}. Figure \ref{fig:conv_lcs_synth} shows the mean light curves produced by the identified candidate solutions. There is good agreement between the candidate light curves and the observations, indicating that high-quality local minima have been identified.

\begin{figure}[H]
  \centering
  \includegraphics[width=\figbig]{converged_solutions_1_Rocket Body.png}
  \caption{Candidate solution light curves found through BFGS optimization.}
  \label{fig:conv_lcs_synth}
\end{figure}

Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} display the candidate MRP and angular velocity of the candidate solutions determined via the procedure in Section \ref{sec:candidate_sols}. In the left Case 1a figures for orientation and angular velocity, we notice that the ground truth in red lies squarely on a surface containing numerous candidate solutions. This indicates that the local minima have correctly converged to the non-singular subspace of states that approximate the observations well for the assumed shape model. In angular velocity space, shown in Figure \ref{fig:omega_sols1}, most solutions fall onto the surface of two open-ended cylinders -- one with half the radius of the larger cylinder containing the truth. Similarly in orientation space, shown in Figure \ref{fig:sigma_sols1}, the solutions fall on two distinct manifolds, one containing the truth -- likely resulting from a reflection of the object's body frame about the bisector between the Sun and observer vectors \cite{marto2024, burton2024journal}. These additional families of solutions highlight the reality that a wide variety of attitude profiles can produce nearly identical observations.

The Case 1b results on the right within Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} show that the majority of the initial states sampled close to the truth have converged to a local subset of the solution manifolds identified by Case 1a. We should not expect these solutions to have shrunk down exponentially towards the ground truth -- the mismatch in the assumed shape model and the measurement noise inherently limits the minimum size of the feasible solution space.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body.png}
    \caption{}
    \label{fig:sigma_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body Prior_pin.png}
    \caption{}
    \label{fig:sigma_sols1b}
  \end{subfigure}

  \caption{Candidate solution initial orientation MRP vectors for Cases 1a (left) and 1b (right), with the ground truth highlighted in red.}
  \label{fig:sigma_sols1}
\end{figure}

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body_pin.png}
    \caption{}
    \label{fig:omega_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body Prior_pin.png}
    \caption{}
    \label{fig:omega_sols1b}
  \end{subfigure}
  \caption{Candidate solution initial body-frame angular velocity vectors for Cases 1a (left) and 1b (right), with the ground truth highlighted in red. TODO: change the perspective on the angular velocities}
  \label{fig:omega_sols1}
\end{figure}

Figure \ref{fig:i_sols1} displays the inertia tensor solutions for Cases 1a and 1b. Notably, the majority of candidate solutions remain within a few standard deviations of the mean of their initial distribution, with the remaining minority exploring directions that maintain the same precession rate. When a good prior for the other attitude state components is available in Case 1b, the inertia tensor solutions are more well-clustered, although due to the low observability of the rocket body's spin rate in the observed light curve, there are still ambiguous solutions with higher or lower $J_z / J_x$ values.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_2_Rocket Body.png}
    \caption{}
    \label{fig:i_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_2_Rocket Body Prior.png}
    \caption{}
    \label{fig:i_sols1b}
  \end{subfigure}
  \caption{Candidate solution inertia ratios for Cases 1a (left) and 1b (right), with the ground truth highlighted in red.}
  \label{fig:i_sols1}
\end{figure}

Figure \ref{fig:w_vs_ang_error_sols1} summarizes the distribution of candidate solutions with respect to the ground truth by displaying the angular velocity magnitude error and angular orientation error compared to the ground truth. Case 1a shows significant variability in angular velocity magnitude, although this is almost entirely due to the unobservability of the spin rate about the object's axis of symmetry in the light curve. The mean precession rate error is low at $0.3\%$. Due to the underdetermined orientation solution space, there are far more solutions at high angular offsets from the truth orientation, leading to a large mean angular error of $126^\circ$. For Case 1b, the mean orientation error is much lower at $18^\circ$, although convergence here is still inherently limited by the noisy measurements and model mismatch.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_3_Rocket Body.png}
    \caption{}
    \label{fig:w_vs_ang_error_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_3_Rocket Body Prior.png}
    \caption{}
    \label{fig:w_vs_ang_error_sols1b}
  \end{subfigure}

  \caption{Candidate solution angular orientation error and angular velocity magnitudes for Cases 1a (left) and 1b (right) compared to the ground truth}
  \label{fig:w_vs_ang_error_sols1}
\end{figure}

% While the state vectors of the candidate solutions do not uniquely indicate the ground truth, the object's precession rate is estimated

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=\figsmall]{spin_proc_Rocket Body.png}
%   \caption{Converged solution body frame spin and precession rates.}
%   \label{fig:spin_proc1}
% \end{figure}

@@

% The two objects selected for study in this work are a Delta I upper stage and the ECHOSTAR 2 satellite.

\subsection{Case 2: Delta I Upper Stage}

>>>>

\usepackage{hyperref}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{times}
\usepackage[numbers]{natbib}
\usepackage{graphicx}
\usepackage{bm} % for bold math
\usepackage{mathtools} % for \prescript
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{adjustbox}

@@

\newcommand{\figbig}[0]{0.4\textwidth}
\newcommand{\figmed}[0]{0.35\textwidth}
\newcommand{\figsmall}[0]{0.3\textwidth}

\newcommand{\subfigbig}[0]{0.4\textwidth}
\newcommand{\subfigsm}[0]{0.29\textwidth}
\newcommand{\subfigmed}[0]{0.43\textwidth}

@@

Methods for estimating the attitude state of uncontrolled space objects from light curves differ significantly depending on the presence of an initial guess. If prior knowledge is available, many filter-based approaches have been developed to process new photometry to update the guess orientation and angular velocity. This class of methods includes single Kalman filters \cite{burton2021two, gagnon2024, wetterer2009}, and multi-filter multiple-model adaptive estimation (MMAE) schemes \cite{linares2014space, dianetti2020}. These approaches can be very effective if the initial state estimate is close to the truth, but have been reported to diverge even in cases when the initial guess is within $5^\circ$ of the ground truth orientation and $0.03^\circ\cdot s^{-1}$ in angular velocity error \cite{gagnon2024} due to measurement nonlinearity and ambiguities.

@@

While much work has performed full state attitude inversion on simulated light curve data \cite{burton2024journal, gagnon2024, robinson2025att}, the applicability of many of these methods to real observations is untested. Studies that do work with real data often focus on estimating an assumed constant spin rate and axis of rotation for the observed object. These assumptions are made to simplify the estimation process but can lead to misleading conclusions when the observed object has a complex rotation state.

@@

The simplifying assumptions of constant, single-axis rotation that are commonly made when working with real data enable elegant solutions but break down when the observed object is in a complex spin profile. Methods for full attitude state solutions -- time-varying orientation and angular velocities -- have been developed \cite{clark2022, burton2024journal, gagnon2024, linares2014particle}, but are seldom applied to real data. Works that have been successful at estimating a full rigid-body attitude state from real data include genetic algorithms \cite{piergentili2020, gallucci2020} and brute-force grid search \cite{shafer2017}. Any full-state estimation method must tackle the inherent computational complexity, lack of knowledge of material reflectivities, and the presence of measurement noise.

In this paper, we describe a global full-state attitude solver inspired by recent PSO work \cite{burton2024journal} but is easier to tune, can be fully parallelized, and is robust to significant measurement noise \cite{robinson2025att}. Further, our method differs from other global estimation methods in that we do not seek to find one optimal solution and instead search for a collection of high-quality local minima to directly study the many ambiguous solutions for a given light curve. Our approach is designed specifically for use with real data as we explicitly take time-varying measurement noise and uncertainties in the body moments of inertia into account.

@@

This work proceeds by describing the inversion method, discussing efficient methods for computing the time-varying probability density of the light curve, and reporting inversion test case results. We present inversion results for a simulated rocket body light curve where the ground truth is available to demonstrate the efficacy of the method, followed by results for real light curve data collected by the Purdue Optical Ground Station for a derelict ECHOSTAR satellite and Delta I rocket body. A deeper treatment of our method and additional simulated light curve results can be found in \cite{robinson2025att}.

@@

We use the Runge-Kutta 4 numerical integration algorithm for all state propagation as it is computationally inexpensive and has low error -- usually below $1^\circ$ over the course of a well-sampled light curve.

@@

\begin{equation} \label{eq:nll_loss}
  \begin{split}
 f(\vctr{x}) = \frac{1}{m}\sum_{k=1}^{m}\Biggl[&\frac{1}{2} \ln2\pi + \ln\sigma_k \\ + &\frac{1}{2}\left(\frac{\vctr{S}_k - \hat{\vctr{S}}_k(\vctr{x})}{\sigma_k}\right)^2 \Biggr].
  \end{split}
 \end{equation}

@@

In this work, initial state vectors are sampled from a prior distribution, usually defined independently for the MRP, angular velocity, and inertia tensor ratios. If no information is available, as is usually the case, a uniform distribution is sampled from the first six components of the state. In particular, $\vctr{p}_{i,0}(0)$ is uniformly randomly sampled from orientation space, $\vctr{\omega}_{i,0}(0)$ is randomly sampled from a uniform distribution of magnitudes up to $\norm{\vctr{\omega}}_\text{mid}$ -- the mean angular rate at which to initialize state samples. We determine this value based on the maximum power positive frequency in the observed light curve $f^\star$ in Hertz as simply $\norm{\vctr{\omega}}_\text{mid} = 2\pi f_\omega$. Uniformly sampling over angular velocity directions, we can express the default sampling strategy for an angular velocity state as:

@@

\noindent
where $g_{1:3} \in \mathcal{N}(0,1)$ and $m=\mathcal{U}(c_1 \norm{\vctr{\omega}}_\text{mid}, c_2 \norm{\vctr{\omega}}_\text{mid})$. The constants $c_1<1$ and $c_2>1$ are chosen to correct for the fact that even for a well-chosen $\norm{\vctr{\omega}}_\text{mid}$, there can still be numerous high-quality solutions at significantly lower or higher angular velocities that possess equal $f^\star$ values. This is a result of object shape symmetries or harmonics in spin and precession rates. We select $c_1=0.5$, $c_2=2$ for this work.

@@

\begin{equation} \label{eq:j_sampler}
  \begin{bmatrix}
 \left(J_y / J_x\right)_{i,0} \\ \left(J_z / J_x\right)_{i,0}
  \end{bmatrix} \sim \mathcal{N} \left(\begin{bmatrix}
 \left(J_y / J_x\right)_\text{est} \\
 \left(J_z / J_x\right)_\text{est} \end{bmatrix}, \sigma_J I_2 \right),
\end{equation}

@@

\subsection{Generating the Simulated Light Curve}

To make this simulated case realistic, we use the presented measurement noise models and introduce a mismatch between the ground truth object used to generate the measurements and the model used for inversion. The selected object is a high-fidelity model of the Saturn V second stage scaled down by a factor $10/3$ to act as a template rocket body object, shown on the right of Figure \ref{fig:rb_models}

@@

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.2} 
  \caption{Simulated observation characteristics}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
 Variable & Value \\ \midrule
 Target COSPAR ID & 1990-054D \\ \grule
 First obs.\ (UTC) & Mar 8, 2025 02:00:00.000 \\ \grule
 Light curve duration & $5$ minutes \\ \grule
 Observations & $100$ \\ \grule
 Integration time & $0.5$ seconds \\ \brule
  \end{tabular}
  \label{tb:case1_in}
\end{table}

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.2} 
  \caption{Reflection properties of materials used in the synthetic data test case. The white paint parameters are estimated qualitatively.}
  \vspace*{6pt}
  \begin{tabular}{@{} l l l l @{}}
    \toprule
 Material & $C_d$ & $C_s$ & $n$ \\ \midrule
 Aluminum \cite{fankhauser2023} & $0.34$ & $0.40$ & $8.9$ \\ \arrayrulecolor{darkgray}\cmidrule(){1-4}
 White paint & $0.9$ & $0.1$ & $1$ \\ \brule
  \end{tabular}
  \label{tb:synth_matprops}
\end{table}

@@

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.2} 
  \caption{True object attitude profile for all test cases}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
 Variable & Value \\ \midrule
 Initial $\vctr{p}(0)$ & $-\frac{1}{3} \begin{bmatrix} 1 & 1 & 1 \end{bmatrix}^T$ \\ \grule
 True $\vctr{\omega}(0)$ ($\text{rad}\cdot\text{s}^{-1}$) & $\vctr{\omega}(0) = \begin{bmatrix} 0.03 & 0.06 & 0.03 \end{bmatrix}^T$ \\ \grule
 Inertia tensor ratios & $J_y / J_x = 1, \: J_z / J_x = 0.25$ \\ \grule
 External torque $\vctr{M}$ & $\begin{bmatrix} 0 & 0 & 0 \end{bmatrix}^T$ \\ \brule
  \end{tabular}
  \label{tb:synth_att}
\end{table}
\FloatBarrier

@@

@@

\subsection{Case 1U and 1P: Synthetic Light Curve Optimization Results}

@@

Two distinct inversion cases are computed. Case 1U presents inversion results given no attitude information besides an estimate of the inertia tensor ratios derived from the known geometry -- hence requiring a uniform search of the solution space. In contrast, Case 1P presents results given good prior knowledge of attitude state distributions. Table \ref{tb:case1Ub_priors} summarizes the initial condition sampling strategies employed for each case. For Case 1P, the standard deviations $\sigma_\omega$ and $\sigma_J$ are for spherically-symmetric Gaussians centered on the ground truth. MRPs are sampled by rotating by $\theta\sim \mathcal{N}(0,\sigma_p)$ degrees away from the ground truth orientation around an axis uniformly sampled on the sphere.

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.2} 
  \caption{Initial condition sampling parameters used for inversion in Cases 1U (no attitude information) and 1P (prior attitude knowledge)}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    \multicolumn{2}{c}{\textbf{Case 1U}} \\
    \midrule
 State & Sampling strategy \\ \midrule
    $\vctr{p}$ & Uniform, via Equation \ref{eq:mrp_sampler} \\ \grule
    $\vctr{\omega}$ & $\norm{\vctr{\omega}}_\text{mid} = 5.26$ $[\text{deg}\cdot \text{s}^{-1}]$, via Eq.\ \ref{eq:omega_sampler} \\ \grule
    $J$ &  $\sigma_J = 0.1$, via Equation \ref{eq:j_sampler} \\ \brule
    \addlinespace
    \toprule
    \multicolumn{2}{c}{\textbf{Case 1P}} \\
    \midrule
 State & Sampling strategy \\ \midrule
    $\vctr{p}$ & Gaussian, $\sigma_p=20$ $[\text{deg}]$ \\ \grule
    $\vctr{\omega}$ & Gaussian, $\sigma_\omega = 0.1$ $[\text{deg}\cdot\text{s}^{-1}]$ \\ \grule
    $J$ & Gaussian, $\sigma_J = 0.05$ $[\text{nondim}]$ \\ \brule
  \end{tabular}
  \label{tb:case1Ub_priors}
\end{table}

For both cases, a total of $n_\text{sample}=10^5$ initial state samples are created via the sampling schemes described in their respective tables. BFGS minimization is then performed for each initial state sample to locally solve the optimization problem given by Equation \ref{eq:opt_problem}. Both cases ran to convergence on an Apple M1 CPU in $3.2$ hours.

Solutions are then classified as candidates for the ground truth with $\ell = 1/2$ via Equation \ref{eq:ell_selection_criteria}. $2026$ and $39683$ candidate solutions were identified for Cases 1U and 1P, respectively. Figure \ref{fig:conv_lcs_synth} shows the mean light curves produced by the identified candidate solutions. There is good agreement between the candidate light curves and the observations, indicating that high-quality local minima have been identified.

\begin{figure}[H]
  \centering
  \includegraphics[width=\figmed]{converged_solutions_1_Rocket Body.png}
  \caption{Candidate solution light curves found through BFGS optimization.}
  \label{fig:conv_lcs_synth}
\end{figure}

Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} display the candidate MRP and angular velocity of the candidate solutions determined via the procedure detailed in Section \ref{sec:candidate_sols}. In the Case 1U figures on the left for orientation and angular velocity, we notice that the ground truth in red lies squarely on a surface containing numerous candidate solutions. This indicates that the local minima have correctly converged to the non-singular subspace of states that approximate the observations well for the assumed shape model. In angular velocity space, shown in Figure \ref{fig:omega_sols1}, most solutions fall onto the surface of two open-ended cylinders -- one with half the radius of the larger cylinder containing the truth. Similarly in orientation space, shown in Figure \ref{fig:sigma_sols1}, the solutions fall on two distinct manifolds, one containing the truth -- likely resulting from a reflection of the object's body frame about the bisector between the Sun and observer vectors \cite{marto2024, burton2024journal}. These additional families of solutions highlight the reality that a wide variety of attitude profiles can produce nearly identical observations.

The Case 1P results on the right within Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} show that the majority of the initial states sampled close to the truth have converged to a local subset of the solution manifolds identified by Case 1U. We should not expect these solutions to have shrunk down exponentially towards the ground truth -- the mismatch in the assumed shape model and the measurement noise inherently limits the minimum size of the feasible solution space.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body.png}
    \caption{}
    \label{fig:sigma_sols1U}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body Prior_pin.png}
    \caption{}
    \label{fig:sigma_sols1P}
  \end{subfigure}

  \caption{Candidate solution initial orientation MRP vectors for Cases 1U (left) and 1P (right), with the ground truth highlighted in red.}
  \label{fig:sigma_sols1}
\end{figure}

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body_pin.png}
    \caption{}
    \label{fig:omega_sols1U}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body Prior_pin.png}
    \caption{}
    \label{fig:omega_sols1P}
  \end{subfigure}
  \caption{Candidate solution initial body-frame angular velocity vectors for Cases 1U (left) and 1P (right), with the ground truth highlighted in red. Note that each plot shows a unique perspective to highlight the structured distribution of solutions.}
  \label{fig:omega_sols1}
\end{figure}

Figure \ref{fig:i_sols1} displays the inertia tensor solutions for Cases 1U and 1P. Notably, the majority of candidate solutions remain within a few standard deviations of the mean of their initial distribution, with the remaining minority exploring directions that maintain the same precession rate. When a good prior for the other attitude state components is available in Case 1P, the inertia tensor solutions are more well-clustered, although due to the low observability of the rocket body's spin rate in the observed light curve, there are still ambiguous solutions with higher or lower $J_z / J_x$ values.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_2_Rocket Body.png}
    \caption{}
    \label{fig:i_sols1U}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_2_Rocket Body Prior.png}
    \caption{}
    \label{fig:i_sols1P}
  \end{subfigure}
  \caption{Candidate solution inertia ratios for Cases 1U (left) and 1P (right), with the ground truth highlighted in red.}
  \label{fig:i_sols1}
\end{figure}

Figure \ref{fig:w_vs_ang_error_sols1} summarizes the distribution of candidate solutions with respect to the ground truth by displaying the angular velocity magnitude error and angular orientation error compared to the ground truth. Case 1U shows significant variability in angular velocity magnitude, although this is almost entirely due to the unobservability of the spin rate about the object's axis of symmetry in the light curve. The mean precession rate error is low at $0.3\%$.

Due to the underdetermined orientation solution space, there are far more solutions at high angular offsets from the truth orientation, leading to a large mean angular error of $126^\circ$. For Case 1P, the mean orientation error is much lower at $18^\circ$, although convergence here is still inherently limited by the noisy measurements and model mismatch.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_3_Rocket Body.png}
    \caption{}
    \label{fig:w_vs_ang_error_sols1U}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{converged_solutions_3_Rocket Body Prior.png}
    \caption{}
    \label{fig:w_vs_ang_error_sols1P}
  \end{subfigure}

  \caption{Candidate solution angular orientation error and angular velocity magnitudes for Cases 1U (left) and 1P (right) compared to the ground truth}
  \label{fig:w_vs_ang_error_sols1}
\end{figure}

While introducing a good prior in Case 1P does constrain the solution space, the ambiguities of the light curve inversion problem are still present and lead to the remaining variability we see in Figure \ref{fig:w_vs_ang_error_sols1P}. Our method respects this fact and resolves the structure of these surfaces, while other approaches that consider only one solution fail.

@@

In this section, we provide inversion results for a Delta I upper stage in Case 2D, and the ECHOSTAR II satellite in Case 2E, both observed by the Purdue Optical Ground Station.

\subsection{Case 2D: Delta I Upper Stage}
